---
title: "Practical Machine Learning Course Project"
author: "Mudit"
date: "Sunday, May 24, 2015"
output: html_document
---

Introduction
----------------------
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 

Data 
-----------------------

The training data for this project are available [here](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv) and test data [here](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv)
The data for this project come from this [source](http://groupware.les.inf.puc-rio.br/har)
After data is stored into working directory, load the data.
```{r}
## read data
training<-read.csv("pml-training.csv",na.strings=c("NA",""))
testing<-read.csv("pml-testing.csv",na.strings=c("NA",""))
```
Library
--------------------------
Load the required library needed for the task.
```{r}
## load required libraries
library(caret)
```
Data Cleaning
--------------------------
In data cleaning following steps were taken:
1. Remove non numeric columns
2. Remove NA
3. Remove near zero variance columns
4. Convert response variable to factor
```{r}
## data cleaning
# cols 1 to 7 (non numeric columns) won't contribute to the model. So remove them.
training   <-training[,-c(1:7)]
# remove NA
training <- training[, which(as.numeric(colSums(is.na(training)))==0)]
# exclude near zero variance features
nearZeroColumns <- nearZeroVar(training, saveMetrics = TRUE)
training <- training[, nearZeroColumns$nzv==FALSE]
#classe into factor
training$classe <- factor(training$classe)
```
Data Partition
--------------------------
Training data is splitted into two parts:
1. Training (75% of original training data)
2. Validation (25% of original training data)
```{r}
## data partition
set.seed(1234)
trainset <- createDataPartition(training$classe, p = 0.75, list = FALSE)
Training <- training[trainset, ]
Validation <- training[-trainset, ]
```
Modelling
--------------------------
Random Forest classifier is used to fit the model. 
```{r}
## Fit a randomforest classifier
library(randomForest)
modelrf<-randomForest(classe ~ ., data = Training,method="class")
```
Training Set Accuracy
--------------------------
```{r}
#predict on train set and check the accuracy
predicttraining <- predict(modelrf, Training)
print(confusionMatrix(predicttraining, Training$classe))
```
Model has performed well in training set, but we will need to cross validate the results against the validation set.

Validation Set Accuracy
--------------------------
```{r}
# validation set accuracy
predictvalidation <- predict(modelrf, Validation)
print(confusionMatrix(predictvalidation, Validation$classe))
```
Cross validation accuracy is 99.61% and the out-of-sample error is therefore 0.39% so the model has performed rather good.

Test set classification
--------------------------
Prediction of our algorithm for the test set:
```{r}
# predict against test set
predicttest <- predict(modelrf, testing)
predicttest
```
Output
---------------
Save the output to files according to instructions and post it to the submission page.
```{r}
answers <- as.vector(predicttest)

pml_write_files = function(x) {
  n = length(x)
  for (i in 1:n) {
    filename = paste0("problem_id_", i, ".txt")
    write.table(x[i], file = filename, quote = FALSE, row.names = FALSE,
                col.names = FALSE)
  }
}

pml_write_files(answers)
```

